RationalGRL: A Framework for Argumentation and Goal Modeling

### 2018-09-15 Response to reviewers2.txt

<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>
<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<< REPLIES FOR REVIEW #2 >>>>>><<<<>>>>>>>>>>>>>>>>>>>>>>
<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>

===== COMMENT #1
In the course of the revision, several of my suggestions for improvement have simply been refused. Hence, I do not consider the improvements sufficient.

===== REPLY #1
We have reconsidered all of your suggestions for improvements in the first revision, and did another cycle of improvements of our paper based on your suggestions. We have replied to those suggestions below.

===== COMMENT #2
The newly added Section 7 is unacceptable from a scientific point of view. First, it is called a "small user evaluation". Then it has a subsection entitled "Experimental setup", but there are no hypotheses. Instead, there is another subsection entitled "Analysis", which starts with "Although our empirical evaluation ...". And why is there another subsection about usability of the tool? If that was the concern, a usability test or a heuristic evaluation would have had to be done.

===== REPLY #2
We have significantly reworked section 7, going through the usual steps for an empirical evaluation and including extra statistical results. The section now includes the hypotheses that have been tested, results, a brief statistical analysis of the results, possible threats to validity and a discussion.

@@

<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>
<<<<<<<<<<<<<<<<<<<<<<< REVISION OF REPLIES FOR REBUTTAL #1 >>>>>>>>>>>>>>>>>>>>>>>>>>>>
<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>

===== ORIGINAL COMMENT #3
I think that the paper should compare the proposed argumentation structure at least with the beliefs in GRL. In general, how does this argumentation approach really differ from one based on beliefs and, in particular, belief revision? Assume that an argument is a belief of some stakeholder revised by another argument, which is a belief of some other stakeholder?

====== REPLY #3
In our first revision, we already added a brief comparison between arguments in RationalGRL and beliefs in GRL in the related work section (under "Beliefs in GRL"). In this revision, we have done two things. First, we have moved the discussion on beliefs in GRL to "related work" under "goal modelling" and further extended and sharpened our discussion and comparison with GRL beliefs. Our main argument, which is now fully in the article, is that in GRL, beliefs are more of an afterthought: they have no formal semantics and it is not possible to further reason with them (i.e. given indicated conflicts between beliefs compute ways in which beliefs can be revised).

Second, we have added a new discussion under “Belief Revision“ in section 8.2 (Future work),  where we discuss belief revision and the possibilities this may offer for future research. A more detailed comparison between RationalGRL and a framework based on belief revision would require us to develop a new belief revision framework connected to GRL and then model, for example, the case study or an extended example using this new belief revision framework for GRL. This would be interesting material for a new paper

@@

===== COMMENT #4 & #5
How would you compare this specific approach with the old gIBIS?
Please discuss why you are trying to have detailed, while simple, rationale, although the recent trend in DR is otherwise?

===== REPLY #4 & #5
In the first revision, we added some discussion in the 2nd paragraph of section 8.1-Design Rationale, noting that a comparison with gIBIS is contained in this discussion (reference [38] includes gIBIS). In this revision, we have made this connection explicit. We have clearly indicated that we combine ideas from DR and goal modeling.

Furthermore, in this revision we have given new reasons (under 8.1, second par, "There are two reasons for this....") for why we connect to goal diagramming methods, even though the trend in DR is to not specifically diagram reasons.

===== ORIGINAL COMMENT #6
I wonder about the formalization in propositional logic. Why is it not used for automated reasoning (at least partly instead of the specific algorithms)?

===== REPLY #6
In the original reply, we indicated that propositional logic in itself does not contain specific algorithms for, for example, INTRO, REPLACE and DISABLE. Furthermore, the algorithms also introduce arguments and attacks, formal notions of which are not in basic propositional logic. The idea is that the formalization is needed for the algorithms, but that just providing a formal semantics of GRL does not directly tell you how to compute/reason with GRL models.

In this revision, we have added a paragraph to the introduction of section 5, where we explain that we are in fact doing automatic reasoning by implementing the argumentation semantics, but that, from a practical point of view, we have chosen to provide algorithms instead of a propositional logic deduction system.

@@

===== COMMENT #7
I am also missing a discussion about the consequences of this binary "nature" of argumentation with respect to argumentation in reality.

===== REPLY #7
In our original reply, we gave references to less "binary" frameworks for argumentation but did not include them in the revised paper. In this revision, we have included a brief discussion on these frameworks (last paragraph of 5.2), and we also refer to work that discusses the relation between "binary" argumentation and argumentation in the real world in more detail (Rahwan et al. 2010, also see the added footnote X).

===== ALL OTHER ORIGINAL COMMENTS WERE SMALLER COMMENTS (STYLE, TYPO, NAMING) WHICH WE ADDRESSED IN THE ORIGINAL REVISION

### 2018-09-15 discussion.tex
Reviewer comments

\paragraph{Design Rationale} Argumentation in software design has been the subject of the research on so-called \emph{design rationale} (DR)~\cite{shum2006hypermedia}, an explicit documentation of the reasons behind decisions made when designing a system or software architecture. DR looks at issues, options and arguments for and against these options in the design of, for example, a software system. Similar to the literature on goal modeling, much of the traditional DR literature provides modeling languages and diagramming tool support for building design rationales. It is in this diagramming functionality that the link with argument diagrams from philosophy, law and AI~\cite{reed2004araucaria,gordon2007visualizing} has been made, where argument diagrams represent reasoning from premises to conclusions. More recent research on DR moves away from the idea that all decision have to be explicitly diagrammed and focuses more on empirically investigating how critical reflection can help when designing \cite{razavian2016two,SchriekEtal2016}, or which parts of the design process are best explicitly documented \cite{falessi2013value}.

Software design and requirements engineering are very closely related \cite{nuseibeh2001weaving} and hence the insights from the DR literature are directly applicable to RE. The RationalGRL framework essentially incorporates the core ideas from DR into goal-oriented requirements engineering by explicitly including arguments for and against the various options into the goal model. Note that we explicitly connect to the extensive literature on goal modeling, and in this respect may seem to go against the trend in DR research to diagram decisions and processes. However, we note that RationalGRL also includes a development process and critical questions that encourage reflection when thinking about the possible goals and functionality of a system, which connects with the more research ideas on DR \cite{razavian2016two,SchriekEtal2016}.

@@

More explicit rationales for goal models are included in the original GRL specification \cite{} in the form of \emph{belief} elements, which can be attached to other intentional elements (goals, tasks) using the different links (positive contribution, negative contribution) of GRL. The content of a belief is a textual description, which can explain, for instance, the rationale for including a goal in a model. In this regard, they seem like arguments or reasons for the different elements in a goal model. However, in the literature the precise usage and semantics of belief elements in GRL remain unclear: according to the metamodel \cite{•} they can be connected to other intentional elements (including beliefs) through any type of link, but in the jUCMNav tool there is a separate \emph{belief link} for connecting beliefs to goals and tasks only, and the beliefs are not used in any of the algorithms to compute the satisfiability of goals and tasks. We are also not aware of many examples of the usage of belief elements in the literature. So what it seems like is that GRL belief elements can be understood as metadata that can be attached to an element

@@

Additionally, from the user evaluation of the tool it followed that in order to implement the tool in realistic settings, other ways may have to be found to incorporate arguments in the goal modeling process. One idea is to use the tool for collaborative on-line goal modeling, similar to Github: since the renaming (i.e. replacement) and deletion (i.e. disabling) of elements are all logged, it is easy for a stakeholder to continue working on a model that was made by another stakeholder. Furthermore, critical questions can be included for other stakeholders to answer at a later date. As a formal underpinning of this asynchronous communication between users, it would make sense to capture requirements engineering and software design processes as dialogs between parties~\cite{finkelstein1989multiparty,BlackEtal2013}, which are a natural fit with the question-answer format employed in the RationalGRL framework. Finally, it would make sense to further develop the RationalGRL Development Process. One idea here is to have stakeholders play a card game (similar to the well-known planning poker) with cards that contain critical, reflective questions (cf. \cite{SchriekEtal2016})

>>>>

\paragraph{Design Rationale} Argumentation in software design has been the subject of the research on so-called \emph{design rationale} (DR), an explicit documentation of the reasons behind decisions made when designing a system or software architecture. Starting already in the 90's with influential issue-driven design methods such as gIBIS and QOC~\cite{shum2006hypermedia},  
 DR looks at issues, options and arguments for and against these options in the design of, for example, a software system. Similar to the literature on goal modeling, much of the traditional DR literature provides modeling languages and diagramming tool support for building design rationales. It is in this diagramming functionality that the link with argument diagrams from philosophy, law and AI~\cite{reed2004araucaria,gordon2007visualizing} has been made, where argument diagrams represent reasoning from premises to conclusions. One of the problems of such methods and tools is the cognitive overload that results from having to learn and use such tools at the same time as having to discuss and think about complex software designs [5]. Furthermore, not much of the research on design reasoning has empirically investigated which techniques and how much information designers use in design reasoning, which information is missing, and how the use of design reasoning techniques affects design discussions and eventually the quality of design discourse. More recent research on DR therefore moves away from the idea that all decision have to be explicitly diagrammed and focuses more on empirically investigating how critical reflection can help when designing \cite{razavian2016two,TangEtal2018}, or which parts of the design process are best explicitly documented \cite{falessi2013value}.

Software design and requirements engineering are very closely related \cite{nuseibeh2001weaving} and hence the insights from the DR literature are directly applicable to RE. The RationalGRL framework essentially incorporates the core ideas from DR into goal-oriented requirements engineering by explicitly including arguments for and against the various options into the goal model. In this respect, we may seem to go against the trend in DR research to focus less on diagramming and more on guidelines for design processes. There are two reasons for this. First, as opposed to DR diagramming, goal modeling is still very much an active area of research \cite{GonccalvesEtal2018} with clear practical applications \cite{NeaceEtal2018}. Second, we note that RationalGRL also includes a development process and critical questions that encourage reflection when thinking about the possible goals and functionality of a system, which connects with the more research ideas on DR \cite{razavian2016two,TangEtal2018}.

@@

More explicit rationales for goal models are included in the original GRL specification \cite{Amyot:2010:EGM:1841349.1841356} in the form of \emph{belief} elements, which can be attached to other intentional elements (goals, tasks) using the different links (positive contribution, negative contribution) of GRL. The content of a belief is a textual description, which can explain, for instance, the rationale for including a goal in a model. In this regard, they seem like arguments or reasons for the different elements in a goal model. However, in the literature the precise usage and semantics of belief elements in GRL remain unclear: according to the metamodel \cite{Amyot:2010:EGM:1841349.1841356} they can be connected to other intentional elements (including beliefs) through any type of link, but in the jUCMNav tool there is a separate \emph{belief link} for connecting beliefs to goals and tasks only, and the beliefs are not used in any of the algorithms to compute the satisfiability of goals and tasks. We are also not aware of many examples of the usage of belief elements in the literature. So what it seems like is that GRL belief elements can be understood as metadata that can be attached to an element

@@

Additionally, from the user evaluation of the tool it followed that in order to implement the tool in realistic settings, other ways may have to be found to incorporate arguments in the goal modeling process. One idea is to use the tool for collaborative on-line goal modeling, similar to Github: since the renaming (i.e. replacement) and deletion (i.e. disabling) of elements are all logged, it is easy for a stakeholder to continue working on a model that was made by another stakeholder. Furthermore, critical questions can be included for other stakeholders to answer at a later date. As a formal underpinning of this asynchronous communication between users, it would make sense to capture requirements engineering and software design processes as dialogs between parties~\cite{finkelstein1989multiparty,BlackEtal2013}, which are a natural fit with the question-answer format employed in the RationalGRL framework. Finally, it would make sense to further develop the RationalGRL Development Process. One idea here is to have stakeholders play a card game (similar to the well-known planning poker) with cards that contain critical, reflective questions (cf. \cite{TangEtal2018}).

### 2018-09-15 formalframework.tex
Reviewer comments

Let us explain these definitions using the example argumentation framework in Figure~\ref{fig:goalmodeling:arg2}, which is a subset of our RationalGRL model from Figure~\ref{fig:example-small3}, containing only arguments $A_4, A_{5},A'_{5},A_{13}$, and $A_{14}$. In this example, there are five admissible sets: $\{A'_{5}\}$, $\{A_{14}\}$, $\{A'_{5},A_{14}\}$, $\{A_4, A_{14}\}$ and $\{A_4, A'_5, A_{14}\}$. In the admissible sets that contain both $A_{4}$ and $A_{14}$, we say that $A_{14}$ defends $A_4$ against its attacker $A_{13}$. Sets containing both $A_{5}$ and $A'_{5}$, or both $A_{13}$ and either $A_4$ or $A_{14}$ are not conflict free. Sets containing $A_5$ are not admissible, as they do not defend $A_5$. Similarly, sets containing $A_4$ but not $A_{14}$ are not admissible as they do not defend themselves against $A_{13}$.

@@

>>>>

Let us explain these definitions using the example argumentation framework in Figure~\ref{fig:goalmodeling:arg2}, which is a subset of our RationalGRL model from Figure~\ref{fig:example-small3}, containing only arguments $A_4, A_{5},A'_{5},A_{13}$, and $A_{14}$. In this example, there are five admissible sets: $\{A'_{5}\}$, $\{A_{14}\}$, $\{A'_{5},A_{14}\}$, $\{A_4, A_{14}\}$ and $\{A_4, A'_5, A_{14}\}$. In the admissible sets that contain both $A_{4}$ and $A_{14}$, we say that $A_{14}$ defends $A_4$ against its attacker $A_{13}$\footnote{In the argumentation literature, argument $A_{14}$ is often said to \emph{reinstate} argument $A_{4}$}. Sets containing both $A_{5}$ and $A'_{5}$, or both $A_{13}$ and either $A_4$ or $A_{14}$ are not conflict free. Sets containing $A_5$ are not admissible, as they do not defend $A_5$. Similarly, sets containing $A_4$ but not $A_{14}$ are not admissible as they do not defend themselves against $A_{13}$.

@@

Note that argument acceptability is essentially a three-valued (or binary if one does not count \emph{undecided}) concept. There are many more fine-grained ways of defining the acceptability of arguments using, for example, probabilities \cite{li2011probabilistic} or counting the number of attacks \cite{grossi2015graded}. However, the ``basic'' argumentation frameworks, and in particular the  have been shown to be close to how humans reason \cite{rahwan2010behavioral}.
